A Large-Scale Study of the Use of Eval in R

### 2020-08-29 inst/paper/main.tex
lost the plot

To represent end-user code in the corpus, we turned to
Kaggle,\footnote{\url{https://www.kaggle.com}} an online platform for
data-science and machine-learning. The website allows people to share
data-science competitions, a data-analysis problem together with data, for
which users try to find the best solution. The solutions, called {kernels},
are then shared back to the platform in the form of either plain scripts or
as notebooks.  One of the most popular competition is about predicting
passenger survival on
Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with \CorpusKaggle
kernels in R (over 1/4 of all available R kernels) which we used for our
corpus.

Unlike CRAN, Kaggle is not a curated repository and therefore there are no
guarantees about the quality of the code. After downloading all of the
\CorpusKaggle kernels and extracting the R code from the various
formats,\footnote{We use rmarkdown package to convert from Jupyer notebook
  to Rmarkdown as well as to convert from Rmarkdown to plain R.} we found
that \CorpusDuplicatedKaggle were whole-file duplicates. From
\CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next
to various runtime exceptions, common problems were parse errors and
misspelled package names.  The final set contains \CorpusFinishedKaggle
kernels with \summaryrnd{CorpusFinishedKaggleCode} lines of R code.

@@

This section presents an overview of the usage of \eval in our corpus.
Richards et al. reported 82\% of JavaScript-enabled web pages using
\eval~\cite{ecoop11}. The case for R is more striking with 100\% of the
programs call the \eval function.\footnote{We combine calls to \c{eval},
  \c{eval.parent}, \c{evalq} and \c{local}.}  This high number is explained
by the fact that the R implementation, its core packages, use \eval
extensively. We differentiate uses based on their origin. A {\bf core} \eval
is one that occurs in the code of a core package.  A {\bf dependent} \eval
occurs in a package imported by the target package. An {\bf own} \eval is
one that occurs in the source code of the target package.

@@

%% SORRY  I HAD TO REFORMAT
\begin{table}[tb] \centering
 \begin{tabular}{r|c|c|c} \hline
Source & Core    & Dependent                      & Own \\\hline
CRAN   & 5.6M &  867K    & 259K \\
       & (1 / 337 / 46K)& (0 / 52 / 58K ) & (0 / 15 / 15K) \\\hline
Kaggle &  1.2M &  470K   & 0\\
 &  (0 / 721 / 18K) & (0 / 290 / 20K) & (0 / 0 / 0)\\\hline
\end{tabular}\caption{Eval calls in CRAN and Kaggle}\label{A}\end{table}

@@

In this section we turn our attention to the expression passed for evaluation to
\eval. The \eval function accepts all types of R objects as inputs, but in the
majority of cases (\AllExpressionInputEvalCallPerc), it is called with
expressions. The remaining \AllValueInputEvalCallPerc are values, \ie, they
evaluate to themselves.
%
%Furthermore, the proportion of values passed to core
%\eval call is \missingTag{CoreEvalValueInputPerc} compared to a significantly higher
%\missingTag{PackageEvalValueInputPerc} for package \eval calls.
%
Passing a value into an eval is effectively an no-op so we look closely at the
expression types the call sites pass to eval across all the runs.
\AllValueInputEvalSitePerc of all call sites always pass values,
\AllExpressionInputEvalSitePerc pass expressions and a very large proportion,
\AllPolymorphicInputEvalSitePerc, pass both expressions and values as inputs
across all program runs. We inspected the eval callsites that accept only values
or both values and expression as inputs. The top three of these call sites
contribute to over ~80\% of all calls to evals. The first one is inside
\c{match.arg} from core. Given a vector of names as strings, this function finds
the default values for parameters with those names in the caller function. Since
the default value for a function parameter can be a complex expression or a
benign value such as \c{NULL}, \c{match.arg} evaluates it through a call to
\eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The second
such \eval call originates from the popular \ggplot package of R which is a DSL
for plotting. \ggplot has it's own object system, \ggproto, which uses \eval
inside the \ggproto function to determine the super-class from which it
inherits. The default value of the superclass is \c{NULL} but in general could
be an arbitrarily complex expression. The third \eval call is in core R \c{str}
function. This function is used to print a summary of R objects. It internally
maps \eval function on the bindings of an environment which results in inputs of
multiples types being passed to \eval.

@@

We looked at the top ten expressions passed to core and package \eval calls. The
most frequent ten expressions to eval calls from core R contribute to 85\% of
all \eval calls.
\begin{table*}[tb]
  \centering
  \begin{tabular}{l|rr} \hline
    Expression & Eval Call & Eval Call \% \\\hline
    \c{c("auto", "shell", "radix")} & 1987105 & 29\%\\
    \c{c("auto", "shell", "quick", "radix")} & 1593169  & 23\%\\
    \c{\{info <- loadingNamespaceInfo(...} & 1008632 &       14\%\\
    \c{c("onLoad", "attach", "detach", "onUnload")}   & 470566 &      6.9\%\\
    \c{c("append", "prepend", "replace")} &              261587&       3.9\% \\
    \c{c("left", "right", "centre", "none")} & 162086     & 2.4\%\\
    \c{c("no", "ifany", "always")}   &                71580 &       1.1\%\\
    \c{c("pearson", "kendall", "spearman")}  & 72962 &      1.1\%\\
    \c{NULL}& 75330  &      1.1\% \\
    \c{Symbol}&                 66279&       1\%\\\hline
  \end{tabular}
  \caption{Top ten eval calls in Core}
  \label{A}
\end{table*}

@@

The most frequent ten expressions to eval calls from CRAN packages contribute to
77.1\% of all \eval calls.
\begin{table*}[tb]
  \centering
  \begin{tabular}{l|rr} \hline
    Expression & Eval Call & Eval Call \% \\\hline
    \c{Environment} &                                                              989302   & 61\%\\
    \c{column[rows] <<- what} &                                                    55677    & 3.5\%\\
    \c{function(value) freduce(value, `_function_list`)} &                         37251    & 2.3\%\\
    \c{NULL} &                         32005    & 2\%\\
    \c{List} &                         22293    & 1.4\%\\
    \c{c("default", "default2012", "default2011" ...}&                             20610    & 1.3\%\\
    \c{force(..1)}                                        &                        20461    & 1.3\%\\
    \c{alist(`_spec`)}                                   &                         18532    & 1.2\%\\
    \c{inner}                                           &                          18530    & 1.2\%\\
    \c{String Vector}                           &                           17487     & 1.1\%\\
  \end{tabular}
  \caption{Top ten eval calls in CRAN}
  \label{A}
\end{table*}

The expression \c{Environment} occurs because of the four callsites
explained above, \c{ggplot2::ggproto}, \c{R6::generator_funs},
\c{future::backtrace} and \c{RModel::str.RMmodel}. The next expression,
\c{column[rows] <<- what}, is used inside the \c{plyr::rbind.fill} function
to merge data frames by assigning concatenated vectors to rows. The \c{<<-}
operator is interesting in that it skips the current scope and assigns in a
parent scope in which the variable is already present. In our corpus, all
these \eval calls contribute to a single side-effect. The expression
\c{function(value) freduce(value, `_function_list`)} arises from the
\c{magrittr::\%>\%} function which is a pipe operator that pipes the output
of previous command to the next one. The expression is evaluated in a custom
environment to create a function binding for evaluating the components of
the pipe.\c{String Vector} and \c{List} also arise from the same function
when a string or a list is piped using the \c{\%>\%} function into the next
expression.  The \c{NULL} arises from \c{R6::generator_funs} function when
the \eval is passed a \c{NULL} argument by the \c{DataMask_generator}
package. The \c{c("default", "default2012", "default2011" ...} pattern
arises from \c{copula::polyG} where it reflectively access the default
expression for its formal parameter and evaluates it. The \c{force(..1)} and
\c{alist(`_spec`)} patterns occur in \c{glue::glue_data} function which
concatenates and interpolates strings. The two patterns occur because the
function captures unevaluated unnamed arguments and maps the evaluation of
\c{force(..1)} on them.  The \c{force} function forces
promises~\cite{oopsla19a} and returns the result of evaluation. The
\c{inner} pattern arises from \c{glue::identity_transformer} which enables
the creation of custom transformation functions for affecting the
interpolation and concatenation of input by the \c{glue} package.

>>>>

To represent end-user code in the corpus, we turned to Kaggle, an online
platform for data-science and machine-learning. The website allows people to
share data-science competitions, a data-analysis problem together with data,
for which users try to find the best solution. The solutions, called
{kernels}, are then shared back to the platform in the form of either plain
scripts or as notebooks.  One of the most popular competition is about
predicting passenger survival on
Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with \CorpusKaggle
kernels in R (over 1/4 of all available R kernels) which we used for our
corpus.

Unlike CRAN, Kaggle is not a curated repository and therefore there are no
guarantees about the quality of the code. After downloading all of the
\CorpusKaggle kernels and extracting the R code from the various
formats,\footnote{We use {\sf rmarkdown} to convert from notebooks to R.} we
found that \CorpusDuplicatedKaggle were whole-file duplicates. From
\CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next
to various runtime exceptions, common problems were parse errors and
misspelled package names.  The final set contains \CorpusFinishedKaggle
kernels with \summaryrnd{CorpusFinishedKaggleCode} lines of R code.

@@

This section presents an overview of the usage of \eval in our corpus.
Richards et al. reported 82\% of JavaScript-enabled web pages using
\eval~\cite{ecoop11}. The case for R is more striking with 100\% of the
programs call the \eval function.\footnote{We combine calls to {\sf eval},
  {\sf eval.parent}, {\sf evalq} and {\sf local}.}  This high number is
explained by the fact that the R implementation, its core packages, use
\eval extensively. We differentiate uses based on their origin. A {\bf core}
\eval is one that occurs in the code of a core package.  A {\bf dependent}
\eval occurs in a package imported by the target package. An {\bf own} \eval
is one that occurs in the source code of the target package.

@@

%% SORRY  I HAD TO REFORMAT
\begin{table}[!h] \centering
 \begin{tabular}{r|c|c|c} \hline
Source & Core    & Dependent                      & Own \\\hline
CRAN   & 5.6M &  867K    & 259K \\
       & (1 / 337 / 46K)& (0 / 52 / 58K ) & (0 / 15 / 15K) \\\hline
Kaggle &  1.2M &  470K   & 0\\
 &  (0 / 721 / 18K) & (0 / 290 / 20K) & (0 / 0 / 0)\\\hline
\end{tabular}\caption{Eval calls in CRAN and Kaggle}\label{A}\end{table}

@@

In this section we turn our attention to the expression passed for
evaluation to \eval. The \eval function accepts all types of R objects as
inputs, but in the majority of cases (\AllExpressionInputEvalCallPerc), it
is called with expressions. The remaining \AllValueInputEvalCallPerc are
values, \ie, they evaluate to themselves.
%
Passing a value into an eval is effectively an no-op so we look closely at
the expression types the call sites pass to eval across all the runs.
\AllValueInputEvalSitePerc of all call sites always pass values,
\AllExpressionInputEvalSitePerc pass expressions and a very large
proportion, \AllPolymorphicInputEvalSitePerc, pass both expressions and
values as inputs across all program runs. We inspected the eval callsites
that accept only values or both values and expression as inputs. The top
three of these call sites contribute to over ~80\% of all calls to
evals. The first one is inside \c{match.arg} from core. Given a vector of
names as strings, this function finds the default values for parameters with
those names in the caller function. Since the default value for a function
parameter can be a complex expression or a benign value such as \c{NULL},
\c{match.arg} evaluates it through a call to \eval. This makes the \eval
inside \c{match.arg} ``polymorhpic''. The second such \eval call originates
from the popular \ggplot package of R which is a DSL for plotting. \ggplot
has it's own object system, \ggproto, which uses \eval inside the \ggproto
function to determine the super-class from which it inherits. The default
value of the superclass is \c{NULL} but in general could be an arbitrarily
complex expression. The third \eval call is in core R \c{str} function. This
function is used to print a summary of R objects. It internally maps \eval
function on the bindings of an environment which results in inputs of
multiples types being passed to \eval.

@@

We looked at the top ten expressions passed to core and package \eval
calls. The most frequent ten expressions to eval calls from core R
contribute to 85\% of all \eval calls.

\begin{table}[!h] \centering
\begin{tabular}{@{}l|rr@{}} \hline
Expression & Eval Call &  \% \\\hline
\c{c("auto", "shell", "radix")} & 198,7105 & 29\%\\
\c{c("auto", "shell", "quick", "radix")} & 1,593,169  & 23\%\\
\c{\{info <- loadingNamespaceInfo(...} & 1,008,632 &       14\%\\
\c{c("onLoad", "attach", "detach", "onUnload")}   & 470,566 &      6.9\%\\
\c{c("append", "prepend", "replace")} &              261,587&       3.9\% \\
\c{c("left", "right", "centre", "none")} & 162,086     & 2.4\%\\
\c{c("no", "ifany", "always")}   &                71,580 &       1.1\%\\
\c{c("pearson", "kendall", "spearman")}  & 72,962 &      1.1\%\\
\c{NULL}& 75330  &      1.1\% \\
\c{Symbol}&                 6,6279&       1\%\\\hline
\end{tabular}
\caption{Top ten eval calls in Core}\label{B}
\end{table}

@@

The most frequent ten expressions to eval calls from CRAN packages
contribute to 77.1\% of all \eval calls.

\begin{table}[!h]  \centering
\begin{tabular}{@{}l@{~}|@{~}r@{~}r@{}} \hline
Expression & Eval Call &  \% \\\hline
\c{Environment} &                                  989302   & 61\%\\
\c{column[rows] <<- what} &                        55677    & 3.5\%\\
\c{function(value) freduce(value, `_function_list`)} & 37251& 2.3\%\\
\c{NULL} &                         32005    & 2\%\\
\c{List} &                         22293    & 1.4\%\\
\c{c("default", "default2012", "default2011" ...}& 20610    & 1.3\%\\
\c{force(..1)}            &                        20461    & 1.3\%\\
\c{alist(`_spec`)}       &                         18532    & 1.2\%\\
\c{inner}               &                          18530    & 1.2\%\\
\c{String Vector}      &                           17487     & 1.1\%\\
\end{tabular}\caption{Top ten eval calls in CRAN} \label{C}
\end{table}

The expression \c{Environment} occurs because of the four callsites
explained above, \c{ggplot2::ggproto}, \c{R6::generator_funs},
\c{future::backtrace} and \c{RModel::str.RMmodel}. The next expression,
\c{column[rows] <<- what}, is used inside the \c{plyr::rbind.fill} function
to merge data frames by assigning concatenated vectors to rows. The \c{<<-}
operator is interesting in that it skips the current scope and assigns in a
parent scope in which the variable is already present. In our corpus, all
these \eval calls contribute to a single side-effect. The expression
\c{function(value) freduce(value, `_function_list`)} arises from the
\c{magrittr::\%>\%} function which is a pipe operator that pipes the output
of previous command to the next one. The expression is evaluated in a custom
environment to create a function binding for evaluating the components of
the pipe.\c{String Vector} and \c{List} also arise from the same function
when a string or a list is piped using the \c{\%>\%} function into the next
expression.  The \c{NULL} arises from \c{R6::generator_funs} function when
the \eval is passed a \c{NULL} argument by the \c{DataMask_generator}
package. The \c{c("default", "default2012", "default2011" ...} pattern
arises from \c{copula::polyG} where it reflectively access the default
expression for its formal parameter and evaluates it. The \c{force(..1)} and
\c{alist(`_spec`)} patterns occur in \c{glue::glue_data} function which
concatenates and interpolates strings. The two patterns occur because the
function captures unevaluated unnamed arguments and maps the evaluation of
\c{force(..1)} on them.  The \c{force} function forces promises and returns
the result of evaluation. The \c{inner} pattern arises from
\c{glue::identity_transformer} which enables the creation of custom
transformation functions for affecting the interpolation and concatenation
of input by the \c{glue} package.
